---
title: "The State of Biking in Madison 2022"
description: |
  A look at several indicators of how many people biked (and took other means of transportation).
author:
  - name: Harald Kliems 
    orcid_id: 0000-0003-2542-0047
    url: https://haraldkliems.netlify.app/
date: 2022-10-12
repository_url: https://github.com/vgXhc/ACS_1y_Madison/
compare_updates_url:
creative_commons: CC BY-SA
categories:
  - Madison (WI)
  - transportation
  - American Community Survey
  - Eco Counter
output: 
  distill::distill_article:
    code_folding: TRUE
    self_contained: false
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


Another year of commuting data from the American Community Survey (ACS) was released. This is an opportunity to look at the state of biking in 2022. Last year, I did a deep dive into the ACS commuting data, as it was the first year that the data was released after the onset of the pandemic. And indeed: Huge changes in commuting, especially an explosion in working from home, could be seen. However, commuting is only one part of our transportation landscape, and this year my review will also look at several other indicators of the state of biking in Madison, as well as making comparisons with some other US cities and national trends.


# Commuting data from ACS: Fewer people work from home; driving and transit usage are up

After an unprecedented rise in working from home between 2019 and 2021, some people returned to the office in 2022. Their share went from over 26% in 2021 to about 20% in 2021. The two modes with the largest increases were driving (58% to 62%; increase within margin of error) and transit (3% to 5%). All other modes didn't change much at all, and the changes were well within the margins of error. Biking ended up at a miserable 2.6% (+/- 0.7%), the lowest percentate in the data reaching back to 2010.
 

```{r, layout="l-page", fig.width=6, fig.asp = 0.618}
library(tmap)
library(sf)
library(tidycensus)
library(tidyverse)
library(gt)
library(ggtext)
library(extrafont)
loadfonts(device = "win", quiet = TRUE) 

library(showtext)

# First argument = name in R
# Second argument = path to .otf-file
font_add('fa-reg', 'fonts/Font Awesome 6 Free-Regular-400.otf')
font_add('fa-brands', 'fonts/Font Awesome 6 Brands-Regular-400.otf')
font_add('fa-solid', 'fonts/Font Awesome 6 Free-Solid-900.otf')
font_add("Arial Narrow", "C:/Windows/Fonts/ARIALN.TTF")
font_add_google("Roboto Condensed", "Roboto Condensed")

# variables <- load_variables(dataset = "acs1/subject", year = "2021")

get_msn_mode_share <- function(year) {
  acs_data <- get_acs(year = year, survey = "acs1", table = "S0801", geography = "place", state = 55, cache_table=T)
  acs_data |> 
    filter(NAME == "Madison city, Wisconsin") |> 
    mutate(year = year)
}


# variable_readable = case_when(
#   variable == "S0801_C01_002" ~ "Drive",
#   variable == "S0801_C01_009" ~ "Transit",
#   variable == "S0801_C01_010" ~ "Walk",
#   variable == "S0801_C01_011" ~ "Bike",
#   variable == "S0801_C01_013" ~ "Work from home",
#   variable == "S0801_C02_002" ~ "Drive, male",
#   variable == "S0801_C02_009" ~ "Transit, male",
#   variable == "S0801_C02_010" ~ "Walk, male",
#   variable == "S0801_C02_011" ~ "Bike, male",
#   variable == "S0801_C02_013" ~ "Work from home, male",
#   variable == "S0801_C03_002" ~ "Drive, female",
#   variable == "S0801_C03_009" ~ "Transit, female",
#   variable == "S0801_C03_010" ~ "Walk, female",
#   variable == "S0801_C03_011" ~ "Bike, female",
#   variable == "S0801_C03_013" ~ "Work from home, female",
#   
#   
# )



msn_mode_share <- map_dfr(c(2010:2019, 2021:2022), get_msn_mode_share)

msn_mode_share <- msn_mode_share |> 
  mutate(gender = case_when(str_detect(variable, "^S0801_C01") ~ "total",
                            str_detect(variable, "^S0801_C02") ~ "male",
                            str_detect(variable, "^S0801_C03") ~ "female"),
         mode_readable = case_when(
           str_detect(variable, "S0801_C0[1-3]_002") ~ "Drive",
           str_detect(variable, "S0801_C0[1-3]_009") ~ "Transit",
           str_detect(variable, "S0801_C0[1-3]_010") ~ "Walk",
           str_detect(variable, "S0801_C0[1-3]_011") ~ "Bike",
           str_detect(variable, "S0801_C0[1-3]_013") ~ "Work from home"))
        
# data frame for the ggrepel labels on the right of the plot
msn_mode_share_2022 <-  msn_mode_share |> 
  filter(year == 2022 & !is.na(mode_readable))

msn_mode_share |> 
  filter(!is.na(mode_readable) & gender == "total") |> 
  group_by(mode_readable, year) |> 
  ggplot(aes(year, estimate, color = mode_readable)) +
  geom_line(linewidth = 1.2) +
  hrbrthemes::scale_color_ipsum(
    #name = element_blank()
    ) +
  geom_crossbar(aes(ymin = estimate - moe, ymax = estimate + moe), alpha = .9,
                  fatten = 1) +
  hrbrthemes::theme_ipsum_rc() +
  scale_x_continuous(breaks = c(2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2021, 2022), minor_breaks = NULL, limits = c(2010, 2023.5)) +
  ylab("estimate (%)") +
  labs(title = "Some Madisonians returned to the office in 2022,\nand they did so by car or bus",
       subtitle ="City of Madison commute mode share and margin of error",
       caption = "American Community Survey 1-year estimates, Table S0801\nVisualization: Harald Kliems") +
    ggrepel::geom_text_repel(data = msn_mode_share_2022 |> filter(gender == "total"), aes(label = paste0(mode_readable, " ", estimate, "%")), nudge_x = 1) +
  theme(legend.position = "none")
```

# How does Madison compare to US-wide trends?

If we look at change over time, it seems like a good idea to ask whether Madison is following a unique path. The following chart compares the Madison in Madison with that of the US as a whole:

```{r}
get_national_mode_share <- function(year) {
  acs_data <- get_acs(year = year, survey = "acs1", table = "S0801", geography = "us", cache_table=T)
  acs_data |> 
    mutate(year = year)
}


# variable_readable = case_when(
#   variable == "S0801_C01_002" ~ "Drive",
#   variable == "S0801_C01_009" ~ "Transit",
#   variable == "S0801_C01_010" ~ "Walk",
#   variable == "S0801_C01_011" ~ "Bike",
#   variable == "S0801_C01_013" ~ "Work from home",
#   variable == "S0801_C02_002" ~ "Drive, male",
#   variable == "S0801_C02_009" ~ "Transit, male",
#   variable == "S0801_C02_010" ~ "Walk, male",
#   variable == "S0801_C02_011" ~ "Bike, male",
#   variable == "S0801_C02_013" ~ "Work from home, male",
#   variable == "S0801_C03_002" ~ "Drive, female",
#   variable == "S0801_C03_009" ~ "Transit, female",
#   variable == "S0801_C03_010" ~ "Walk, female",
#   variable == "S0801_C03_011" ~ "Bike, female",
#   variable == "S0801_C03_013" ~ "Work from home, female",
#   
#   
# )



national_mode_share <- map_dfr(c(2010:2019, 2021:2022), get_national_mode_share)

national_mode_share <- national_mode_share |> 
  mutate(gender = case_when(str_detect(variable, "^S0801_C01") ~ "total",
                            str_detect(variable, "^S0801_C02") ~ "male",
                            str_detect(variable, "^S0801_C03") ~ "female"),
         mode_readable = case_when(
           str_detect(variable, "S0801_C0[1-3]_002") ~ "Drive",
           str_detect(variable, "S0801_C0[1-3]_009") ~ "Transit",
           str_detect(variable, "S0801_C0[1-3]_010") ~ "Walk",
           str_detect(variable, "S0801_C0[1-3]_011") ~ "Bike",
           str_detect(variable, "S0801_C0[1-3]_013") ~ "Work from home"))
        
# data frame for the ggrepel labels on the right of the plot
national_mode_share_2022 <-  national_mode_share |> 
  filter(year == 2022 & !is.na(mode_readable))
showtext_auto()
national_mode_share |> 
  filter(!is.na(mode_readable) & gender == "total") |> 
  group_by(mode_readable, year) |> 
  ggplot(aes(year, estimate, color = mode_readable)) +
  geom_line(size = 1.2) +
  hrbrthemes::scale_color_ipsum(
    #name = element_blank()
    ) +
  geom_crossbar(aes(ymin = estimate - moe, ymax = estimate + moe), alpha = .9,
                  fatten = 1) +
  hrbrthemes::theme_ipsum_rc() +
  scale_x_continuous(breaks = c(2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2021, 2022), minor_breaks = NULL, limits = c(2010, 2023.5)) +
  ylab("estimate (%)") +
  labs(title = "Some Madisonians returned to the office in 2022,\nand they did so by car or bus",
       subtitle ="City of Madison commute mode share, 2010-2022",
       caption = "American Community Survey 1-year estimates, Table S0801\nVisualization: Harald Kliems") +
    ggrepel::geom_text_repel(data = national_mode_share_2022 |> filter(gender == "total"), aes(label = paste0(mode_readable, " ", estimate, "%")), nudge_x = 1) 


combined <- rbind(national_mode_share |> mutate(location = "national"), msn_mode_share |> mutate(location = "Madison"))
combined_2022 <- rbind(national_mode_share_2022 |> mutate(location = "national"), msn_mode_share_2022 |> mutate(location = "Madison"))

combined |> 
  filter(!is.na(mode_readable) & gender == "total") |> 
  group_by(mode_readable, year, location) |> 
  ggplot(aes(year, estimate, color = mode_readable)) +
  geom_line(size = 1.2) +
  hrbrthemes::scale_color_ipsum(
    #name = element_blank()
    ) +
  # geom_crossbar(aes(ymin = estimate - moe, ymax = estimate + moe), alpha = .9, fatten = 1) +
  hrbrthemes::theme_ipsum() +
  scale_x_continuous(breaks = c(2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2021, 2022), minor_breaks = NULL, limits = c(2010, 2023.5)) +
  ylab("estimate (%)") +
  labs(title = "Mode share trends in Madison follow\na similar pattern as the national numbers",
       subtitle ="City of Madison and US commute mode share, 2010-2022",
       caption = "American Community Survey 1-year estimates, Table S0801\nVisualization: Harald Kliems") +
    ggrepel::geom_text_repel(data = combined_2022 |> filter(gender == "total"), aes(label = paste0(mode_readable, " ", estimate, "%")), nudge_x = 1) +
  theme(legend.position = "none", plot.title = element_text(lineheight = .5, margin = margin(b = 0.5)), plot.subtitle = element_text(margin = margin(t = 0))) +
  facet_wrap(~location)
```
A similar pattern emerges. While Madison commuters drive less, bike, walk, and bus more than the average US commuter, the trend over time looks similar. And that's true for the change from 2021 to 2022 as well: Just like in Madison, nationwide working from home went down, and driving and transit rebounded slightly. 

```{r}
get_place_mode_share <- function(year, places) {
  acs_data <- get_acs(year = year, survey = "acs1", table = "S0801", geography = "place", cache_table=T)
  acs_data |> 
    filter(NAME %in% places) |> 
    mutate(year = year)
}

national_mode_share_2022_total <- national_mode_share_2022 %>% 
  filter(gender == "total") %>% 
  mutate(
    label = case_when(
           mode_readable == "Drive" ~ "<span style='font-family:fa-solid;'>&#xf1b9;</span>",
           mode_readable == "Transit" ~ "<span style='font-family:fa-solid;'>&#xf207;</span>",
           mode_readable == "Walk" ~ "<span style='font-family:fa-solid;'>&#xf554;</span>",
           mode_readable == "Bike" ~ "<span style='font-family:fa-solid;'>&#xf206;</span>",
           mode_readable == "Work from home" ~ "<span style='font-family:fa-solid;'>&#xe066;</span>"
  ),
  NAME = "US")

mode_share_comparison <- get_place_mode_share(2022, c("Madison city, Wisconsin", "Portland city, Oregon", "Minneapolis city, Minnesota", "San Francisco city, California", "Milwaukee city, Wisconsin", "New York city, New York"))

mode_share_comparison <- mode_share_comparison |> 
  mutate(gender = case_when(str_detect(variable, "^S0801_C01") ~ "total",
                            str_detect(variable, "^S0801_C02") ~ "male",
                            str_detect(variable, "^S0801_C03") ~ "female"),
         mode_readable = case_when(
           str_detect(variable, "S0801_C0[1-3]_002") ~ "Drive",
           str_detect(variable, "S0801_C0[1-3]_009") ~ "Transit",
           str_detect(variable, "S0801_C0[1-3]_010") ~ "Walk",
           str_detect(variable, "S0801_C0[1-3]_011") ~ "Bike",
           str_detect(variable, "S0801_C0[1-3]_013") ~ "Work from home"),
         label = case_when(
           mode_readable == "Drive" ~ "<span style='font-family:fa-solid;'>&#xf1b9;</span>",
           mode_readable == "Transit" ~ "<span style='font-family:fa-solid;'>&#xf207;</span>",
           mode_readable == "Walk" ~ "<span style='font-family:fa-solid;'>&#xf554;</span>",
           mode_readable == "Bike" ~ "<span style='font-family:fa-solid;'>&#xf206;</span>",
           mode_readable == "Work from home" ~ "<span style='font-family:fa-solid;'>&#xe066;</span>"
         ),
         NAME = str_remove(NAME, " city")) |> 
    filter(!is.na(mode_readable) & gender == "total")

```

```{r}
# showtext_auto(enable = F)
# detach(extrafont, unload = T)
# detach(ggtext, unload = T)

p <- rbind(mode_share_comparison, national_mode_share_2022_total) %>% 
  mutate(name_by_bike_share = fct_relevel(NAME, c("San Francisco, California", 
                        "Portland, Oregon",
                        "Madison, Wisconsin", 
                        "Minneapolis, Minnesota", 
                        "New York, New York",
                        "Milwaukee, Wisconsin",
                        "US"))) %>% 
  group_by(mode_readable, name_by_bike_share) |> 
  ggplot(aes(name_by_bike_share, estimate, color = mode_readable, label = label)) +
  
    geom_point(size = 3, color = "darkgrey") +
  hrbrthemes::scale_color_ipsum() +
  hrbrthemes::theme_ipsum() +
  # scale_x_continuous(breaks = c(2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2021, 2022), minor_breaks = NULL, limits = c(2010, 2023.5)) +
  ylab("estimate (%)") +
  xlab(element_blank()) +
  labs(title = "2022 Commute mode share",
       subtitle = "Madison and select US cities",
       caption = "American Community Survey 1-year estimates, Table S0801\nVisualization: Harald Kliems") +
    # ggrepel::geom_text_repel(data = combined_2022 |> filter(gender == "total"), aes(label = paste0(mode_readable, " ", estimate, "%")), nudge_x = 1) +
  theme(legend.position = "none",
        panel.grid.major.y = element_blank()) +
  coord_flip()
  # facet_wrap(~NAME)
library(showtext)

showtext_auto(enable = T)
p +
  geom_richtext(size = 8, label.colour = NA, fill = NA)
showtext_auto(enable = F)
```

From 2019 to 2021, the share of people working from home almost quintupled! Fewer people drove to work, and commuting by bus became much less common. For walking and biking the changes were downward too but within the margins of error.

```{r, layout="l-page", fig.width=7, fig.asp = 0.618}
msn_mode_share |> 
  filter(!is.na(mode_readable) & gender == "total" & year >= 2019) |> 
  group_by(mode_readable, year) |> 
  pivot_wider(names_from = year, values_from = c(estimate, moe)) |> 
  mutate(diff_22_21 = estimate_2022-estimate_2021,
         diff_22_21_formatted = paste0(round(diff_22_21,digits = 1), "%")) |>
  ggplot() +
    geom_segment(aes(x=mode_readable, xend=mode_readable, y=estimate_2021, yend=estimate_2022), color="grey", arrow = arrow(length = unit(2, "mm"))) +
  geom_point(aes(x = mode_readable, y = estimate_2021), color = "#d18975") +
    geom_crossbar(aes(mode_readable, estimate_2021, ymin = estimate_2021 - moe_2021, ymax = estimate_2021 + moe_2021), alpha = .9,
                  fatten = 1, color = "#d18975")+
    geom_point(aes(x = mode_readable, y = estimate_2022), color = "#8fd175") +
      geom_crossbar(aes(mode_readable, estimate_2022, ymin = estimate_2022 - moe_2022, ymax = estimate_2022 + moe_2022), alpha = .9,
                  fatten = 1, color = "#8fd175")+
  geom_text(aes(x = mode_readable, 
                y = (estimate_2022 + estimate_2021)/2, 
                label = diff_22_21_formatted),
            nudge_x = .25)+
  hrbrthemes::scale_color_ipsum(
    #name = element_blank()
    ) +
  hrbrthemes::theme_ipsum() +
  coord_flip() +
  theme(panel.grid.major.y = element_blank()) +
  ylab("estimate (%)") +
  xlab(element_blank()) +
  labs(title = "Working from home is down; transit commutes are up",
       subtitle ="City of Madison, 2021-2022. Boxes show margin of error",
       caption = "American Community Survey 1-year estimates, Table S0801\nVisualization: Harald Kliems") 
```

For transportation planning, it's important to not only look at percentages but also at absolute numbers: Fewer people driving to work means less peak hour congestion (which is not [necessarily a good thing](https://ssti.us/2021/03/16/traffic-congestion-vanished-and-more-people-died-on-the-roads/)); fewer transit riders means lower fare revenues; and fewer bike commuters may lead to less support of biking infrastructure. And a large number of people working from home may be a problem for commercial real estate or fewer customers for downtown restaurants. 

```{r, layout="l-page", , fig.width = 8, fig.asp = 0.618}
get_commute_total <- function(year) {
  get_acs(geography = "place",
                           state = "WI",
                           survey = "acs1",
                           year = year,
                           table = "C08006",
          summary_var = "C08006_001",
          cache_table = T) |> 
    mutate(year = year,
           )
}


commute_numbers <- map_dfr(c(2021:2022), get_commute_total)

commute_numbers_msn <- commute_numbers |> 
  filter(NAME == "Madison city, Wisconsin") |> 
  mutate(mode_readable = case_when(
           str_detect(variable, "C08006_001") ~ "Total",
           str_detect(variable, "C08006_002") ~ "Drive",
           str_detect(variable, "C08006_008") ~ "Transit",
           str_detect(variable, "C08006_009") ~ "Bike",
           str_detect(variable, "C08006_010") ~ "Walk",
           str_detect(variable, "C08006_012") ~ "Work from home")) |> 
  filter(!is.na(mode_readable))

commute_numbers_msn |> 
  pivot_wider(id_cols = mode_readable, names_from = year, values_from = c(estimate, moe)) |> 
  mutate(mode_readable = fct_reorder(mode_readable, estimate_2022)) |> 
  group_by(mode_readable) |> 
  mutate(moe_diff = moe_sum(moe = c(moe_2022, moe_2021), estimate = c(estimate_2022, estimate_2021))) |> 
  ggplot() +
    geom_segment(aes(x=mode_readable, xend=mode_readable, y=estimate_2021, yend=estimate_2022), color="grey", arrow = arrow(length = unit(2, "mm"))) +
  geom_point(aes(x = mode_readable, y = estimate_2021), color = 2021) +
    geom_point(aes(x = mode_readable, y = estimate_2022), color = 2022) +
  geom_text(aes(x = mode_readable, 
                y = (estimate_2022 + estimate_2021)/2, 
                label = paste0(scales::number_format(style_positive = "plus",big.mark = ",")(estimate_2022-estimate_2021), " (±", round(moe_diff, 0), ")")),
            nudge_x = .25)+
  hrbrthemes::scale_color_ipsum(
    #name = element_blank()
    ) +
  hrbrthemes::theme_ipsum() +
  coord_flip() +
  theme(panel.grid.major.y = element_blank()) +
  ylab("number of workers") +
  xlab(element_blank()) +
  labs(title = "Change in number of commuters, 2021-2022",
       subtitle ="City of Madison. Statistically insignifant changes greyed out",
       caption = "American Community Survey 1-year estimates, Table C08006\nVisualization: Harald Kliems") +
  gghighlight::gghighlight(abs(estimate_2022-estimate_2021) > moe_diff, use_direct_label = F)
```

The total number of workers and of bike and walk commuters didn't change significantly. But there are about 21000 fewer drivers, 11,000 fewer bus commuters, and over 30,000 more people working from home. For a city of Madison's size, these changes are big!

## What about those still going to the office?

When I first shared the above graphs some people pointed out: "Working from home isn't really commuting! What do these percentages look like when we only account for the people who still leave the house to get to work?" Or to put it differently: Before the pandemic, you and 99 of your coworkers went to the office every day, and 10 of you biked. In 2021, half of your coworkers now work from home. Of the remaining 50 people in the office, do you still have 5 (i.e. 10%) people who bike to work? 

```{r, layout="l-page", fig.width=6, fig.asp = 1}
get_msn_mode_counts <- function(year) {
  acs_data <- get_acs(year = year, survey = "acs1", table = "C08006", geography = "place", state = 55, cache_table=T, summary_var = "C08006_001")
  acs_data |> 
    filter(NAME == "Madison city, Wisconsin") |> 
    mutate(year = year)
}

msn_counts <- map_dfr(c(2021, 2022), get_msn_mode_counts)

msn_counts <- msn_counts |> 
  mutate(summary_est = case_when(year == 2021 ~summary_est - 40279,
                                 year == 2022 ~ summary_est - 31368),
         mode_readable = case_when(
    variable == "C08006_001" ~ "Total",
      variable == "C08006_002" ~ "Drive",
      variable == "C08006_003" ~ "Drove alone",
      variable == "C08006_004" ~ "Carpooled",
      variable == "C08006_008" ~ "Transit",
      variable == "C08006_009" ~ "Bike",
      variable == "C08006_010" ~ "Walk",
      variable == "C08006_011" ~ "Other",
      variable == "C08006_012" ~ "Work from home"
  ),
  mode_share = estimate/summary_est) |> 
  filter(!is.na(mode_readable)) |> 
  filter(mode_readable %in% c("Drive", "Transit", "Bike", "Walk", "Other")) |> 
  mutate(mode_readable = fct_relevel(mode_readable, 
                                     "Other",
                                                "Bike", 
                                                "Walk", 
                                                "Transit", 
                                                "Drive")) |> 
  group_by(year,mode_readable) %>% 
    mutate(pos = cumsum(mode_share) - mode_share/2)
msn_counts |> 
  ggplot(aes(as.factor(year), mode_share, fill = mode_readable)) +
  geom_col(position = "stack") +
  geom_text(aes(label = mode_readable), position = position_stack(vjust = 0.5), color = "white",
            size = 3) +
  scale_y_continuous(labels = scales::label_percent()) +
    hrbrthemes::scale_fill_ipsum(name = "element_blank()") +
  hrbrthemes::theme_ipsum() +
  ylab("Commute mode share estimate") +
  xlab(element_blank()) +
  theme(legend.position = "none",   panel.grid.major.x = element_blank()) +
  labs(title = "xx",
       subtitle ="City of Madison, working from home excluded",
       caption = "American Community Survey 1-year estimates, Table B08006\nVisualization: Harald Kliems")
```

This is the same data in table form.

```{r}
msn_counts |> 
  
  pivot_wider(id_cols = mode_readable, names_from = year, values_from = mode_share, names_prefix = "mode_share_") |> 
  mutate(change = mode_share_2022- mode_share_2021) |> 
  select(mode_readable, mode_share_2021, mode_share_2022, change) |>
  ungroup() |> 
  gt() |> 
  tab_header(title = "Change in commute mode for workers not working from home") |> 
  fmt_percent(columns = c(mode_share_2022, mode_share_2021, change), decimals = 0) |> 
  tab_spanner(
    label = "Mode share",
    columns = c(mode_share_2021, mode_share_2022)
  ) |> 
  cols_label(mode_readable = "Commute mode", 
             mode_share_2022 = "2022",
             mode_share_2021 = "2021",
             change = "Change (percentage points)") |> 
   data_color(
    columns = c(change),
    colors = scales::col_numeric(
      palette = "viridis",
    domain = NULL )) |> 
   tab_source_note(
    source_note = "Data: American Community Survey 1-year estimates, Table B08006. Margins of error not shown."
  )
```

Calculating the margins of error for these estimates is complex and I have not done that. But similar to the error bars we have seen above, changes of less than 2% points are probably not meaningful. So what do we take away from the table and chart? Of the people who still commute to the office, taking the bus is much less common while driving has gone up. The other percentages are basically unchanged. This makes sense: Bus service was reduced and many people perceived riding the bus as a risk for infection. Conversely, peak hour congestion for drivers went down (remember: overall the number of people commuting by car went _down_ even when their percentage here went _up_) and some employers reduced or waived parking fees.





# Counter data 2022
```{r}
counts_2022 <- readxl::read_excel("data/EcoCounter_2022.xlsx", skip = 3,
                   col_names = c("time_count", "count_cap_city", "count_sw_path")) |> 
  mutate(date_count = floor_date(time_count, unit = "hours")) |> 
  summarize(across(starts_with("count_"), ~ sum(.x, na.rm = T)), .by = date_count) |> 
  pivot_longer(cols = starts_with("count_"), names_to = "location", values_to = "count_hourly") |> 
  mutate(location = case_when(location == "count_cap_city" ~ "Cap City at North Shore",
                              location == "count_sw_path" ~ "SW Path at Randall"),
         dayofweek = wday(date_count),
         weekendind = ifelse(dayofweek %in% c(1:5), "weekday", "weekend"),
         month_count = month(date_count, label = T, abbr = T)) 
counts_2022 |>   
group_by(location, month_count) |> 
    summarize(count_monthly = sum(count_hourly)) |> 
  ggplot(aes(month_count, count_monthly, fill = location)) +
  geom_col(position = "dodge") 
    
```

```{r}
# code reused from https://github.com/vgXhc/madison_bike_counts/blob/master/analysis.R

##get bike counter data
cc_counts <- read_csv("data/Eco-Totem_Capital_City_Trail_Bike_Counts(3).csv", col_types = "ci-") %>% mutate(location = "Cap City at North Shore")
sw_counts <- read_csv("data/Eco-Totem_Southwest_Path_Bike_Counts(2).csv", col_types = "ci-") %>% mutate(location = "SW Path at Randall")
#combine two counter locations
counts <- bind_rows(cc_counts, sw_counts)
#some data prep for counts
counts2 <- counts %>% 
  drop_na %>% 
  mutate(date_count = mdy_hm(Count_Date), #fix date and time
         location = as.factor(location),
         # Count = ifelse(Count == 0, 1, Count), #convert 0 counts to 1 to allow log transform
         # log_count = log(Count), #create value for log of count
         dayofweek = wday(date_count),
         weekendind = ifelse(dayofweek %in% c(1:5), "weekday", "weekend"),
         month_count = month(date_count, label = T, abbr = T)
         ) |> 
  select(-Count_Date) |> 
  rename(count_hourly = Count) |> 
  filter(date_count < ymd_hms("2022-01-01 00:00:00"))


rbind(counts2, counts_2022) |> 
  mutate(year_count = year(date_count)) |> 
  filter(year_count >= 2016) |> 
  summarize(count_annual_by_location = sum(count_hourly), .by = c(location, year_count)) |>
  reframe(location, year_count, count_annual_by_location, count_annual = sum(count_annual_by_location), .by = year_count) |> 
  ggplot(aes(year_count, count_annual_by_location, fill = location)) +
  geom_col(position = "stack") +
  geom_text(
    aes(label = after_stat(y), group = year_count), 
    stat = 'summary', fun = sum, vjust = -1
  )
```


# Job growth

```{r}
library(lehdr)
library(sf)
dane_county <- tigris::block_groups(55, 25, cb = T)
jobs_2019 <- grab_lodes(state = "WI", 
                   year = 2019, 
                   lodes_type = "wac", 
                   agg_geo = "bg",
                   use_cache = TRUE)

jobs_2015 <- grab_lodes(state = "WI", 
                   year = 2015, 
                   lodes_type = "wac", 
                   agg_geo = "bg",
                   use_cache = TRUE)

jobs_change <- rbind(jobs_2015, jobs_2019) |> 
  select(year, C000, w_bg) |> 
  pivot_wider(names_from = year, 
              values_from = C000,
              names_prefix = "jobs_") |> 
  mutate(job_growth = jobs_2019 - jobs_2015) 

jobs_change <- dane_county |> 
  left_join(jobs_change, by = join_by(GEOID == w_bg)) |> 
  filter(ALAND != 0) 

tm_shape(jobs_change) +
  tm_polygons("job_growth", style = "jenks",
              alpha = .7)


jobs_dane <- dane_county |> 
  left_join(jobs, by = join_by(GEOID == w_bg)) |> 
  filter(ALAND != 0) |> 
  mutate(jobs_per_acre = C000/(ALAND*0.0002471054))

library(tmap)
tmap_mode("view")
tm_shape(jobs_dane) +
  tm_polygons(col = "jobs_per_acre", style = "jenks",
              alpha = .7)

```

Could try comparing job growth with BNA score
 - maybe not available at BG level? Maybe not available beyond city limits?
 
could also run analysis with distance from capitol (43.074722, -89.384167) -- that'd be fun

# BNA
```{r}
bna_blocks <- st_read("data/neighborhood_census_blocks/neighborhood_census_blocks.shp")


bna_blocks_geoid <- bna_blocks |> 
  mutate(block_group_geoid = str_trunc(BLOCKID10, 
                                       width = 12, 
                                       side = "right",
                                       ellipsis = ""
                                      )) |> 
  summarize(bna_score_bg = mean(OVERALL_SC, na.rm = T), .by = block_group_geoid)
jobs_change |> 
  left_j
  
bna_blocks_geoid |> st_drop_geometry() |> 
  left_join(dane_county, by = join_by(block_group_geoid == GEOID)) |> class()
  tm_shape() +
  tm_polygons("bna_score_bg")
bna_ways <- st_read("data/neighborhood_ways/neighborhood_ways.shp")
tmap_mode("view")
bna_blocks |> 
  tm_shape() +
  tm_polygons("OVERALL_SC")
```




